---
title: When Does Sorting Matter
author: Roel M. Hogervorst
date: '2020-05-07'
slug: when-does-sorting-matter
categories:
  - R
tags:
  - ggplot2
  - sprite
output: 
  html_document:
    keep_md: true
---

I was working on an implementation of SPRITE [(Sample Parameter Reconstruction  via Iterative TEchniques)](https:://peerj.preprints.26968v1 "Heathers JA, Anaya J, van der Zee T, Brown NJ. 2018. Recovering data from summary statistics: Sample Parameter Reconstruction via Iterative TEchniques (SPRITE) PeerJ Preprints 6:e26968v1 https://doi.org/10.7287/peerj.preprints.26968v1")
and looking into possible speedups of the process. 

```{r setup, include=FALSE}
# Options to have images saved in the post folder
# And to disable symbols before output
knitr::opts_chunk$set(fig.path = "", comment = "")

# knitr hook to make images output use Hugo options
knitr::knit_hooks$set(
  plot = function(x, options) {
    hugoopts <- options$hugoopts
    paste0(
      "{{<figure src=",
      '"', x, '" ',
      if (!is.null(hugoopts)) {
        glue::glue_collapse(
          glue::glue('{names(hugoopts)}="{hugoopts}"'),
          sep = " "
        )
      },
      ">}}\n"
    )
  }
)

# knitr hook to use Hugo highlighting options
knitr::knit_hooks$set(
  source = function(x, options) {
  hlopts <- options$hlopts
    paste0(
      "```r ",
      if (!is.null(hlopts)) {
      paste0("{",
        glue::glue_collapse(
          glue::glue('{names(hlopts)}={hlopts}'),
          sep = ","
        ), "}"
        )
      },
      "\n", glue::glue_collapse(x, sep = "\n"), "\n```\n"
    )
  }
)
```

In general I think there are 2 ways to speed up processes:

- Do the same thing, but faster (using a different language, using vectorization, pre-allocating structures)
- Do the thing smarter (skip steps, use different structures)


Sprite is a tool to recreate underlying data when you only have the summary statistics. 
The sprite process creates a vector of integers of the size as the real sample size and 
tweaks it by incrementing and or decrementing values in that vector. 


One of the possible speedups would be to use sorting. To identify candidate values in the vector the process does a lot of vector operations which are quite fast, but that does mean that for 
every operation the process needs to check
every value in the vector for a condition (is it bigger than x, is it the max, etc.). 

Theoretically, if the 
vector was sorted, the process would be quicker (if this value is bigger, than all values after
are also, etc.). But the sorting process takes time too. Is the time win for these operations 
enough for the extra cost of sorting the vector?

Of course! I hear you think. But most of the experiments in social and medical sciences, the primary target for this tool, are smaller than 1000 samples.

Let's simulate:

```{r packages needed}
library(tidyverse) # I want dplyr, ggplot2 and some string manipulation too
library(microbenchmark)
```

Some functions:

```{r}
# create unsorted integer representation. of length 10^ c(1, 2,3,4,5,6,7,8,9)
# 9 is already too much!
# more doesn't matter for this purpose.
create_int_vector_length_n <- function(n){
    scale_max = 31L  # using int here makes it go from 5someting to 4somthing.
    scale_min = 1L
    mean = 15
    pmax(pmin(as.integer(runif(n) * 2 * mean), scale_max), scale_min)
}
# additive cost of sorting
create_and_sort_n<- function(n){
    vec <- create_int_vector_length_n(n)
    sort(vec)
}
#### actions (it is really not more sophisticated than this)
# smaller than max
# smaller than value
# duplicate
# not equal to a value
execute_functions <- function(vec){
    endresult = !duplicated(vec) & !(vec %in% c(2,3)) & vec < 11
}
```

Then we can check the cost of sorting.

```{r}
l1= create_int_vector_length_n(10^1)
l2= create_int_vector_length_n(10^2)
l3= create_int_vector_length_n(10^3)
l4= create_int_vector_length_n(10^4)
l5= create_int_vector_length_n(10^5)
l6= create_int_vector_length_n(10^6)

sorting_only <- microbenchmark(
    l1=sort(l1), l2=sort(l2),
    l3= sort(l3), l4 = sort(l4), l5 = sort(l5), l6= sort(l6)
)
ggplot2::autoplot(sorting_only) + ggtitle("Integer vector sorting")
```

Yes this is measured in microseconds, sorting an integer vector of length 1000 takes at least 0.1 second.

How long does it take to create a vector?

```{r}
creation <- microbenchmark(
    l1= create_int_vector_length_n(10^1), l2= create_int_vector_length_n(10^2),
    l3= create_int_vector_length_n(10^3), l4= create_int_vector_length_n(10^4),
    l5= create_int_vector_length_n(10^5), l6= create_int_vector_length_n(10^6)
)
ggplot2::autoplot(creation)+ggtitle("vector creation time")
```

Creation takes about as much time for 1000 to 10.000 samples.

So are the functions faster on sorted data?

```{r}
l1_sorted <- sort(l1)
l2_sorted <- sort(l2)
l3_sorted <- sort(l3)
l4_sorted <- sort(l4)
l5_sorted <- sort(l5)
l6_sorted <- sort(l6)
sort_vs_unsorted <- microbenchmark(
    l1=execute_functions(l1),
    l1_sort = execute_functions(l1_sorted),
    l2=execute_functions(l2),
    l2_sort = execute_functions(l2_sorted),
    l3= execute_functions(l3),
    l3_sort = execute_functions(l3_sorted),
    l4= execute_functions(l4),
    l4_sort = execute_functions(l4_sorted),
    l5= execute_functions(l5),
    l5_sort = execute_functions(l5_sorted),
    l6= execute_functions(l6),
    l6_sort = execute_functions(l6_sorted)
)
ggplot2::autoplot(sort_vs_unsorted)+
    labs(title="Sorting does not seem to help us")
```

Let's do a more fair comparison

```{r}
sort_and_execute_functions<- function(vec){
    vec <- sort(vec)
    endresult <- execute_functions(vec)
}
comparison <- microbenchmark(
    l1 = execute_functions(l1),
    l1_s = sort_and_execute_functions(l1),
    l2 = execute_functions(l2),
    l2_s = sort_and_execute_functions(l2),
    l3 = execute_functions(l3),
    l3_s = sort_and_execute_functions(l3),
    l4 = execute_functions(l4),
    l4_s = sort_and_execute_functions(l4),
    l5 = execute_functions(l5),
    l5_s = sort_and_execute_functions(l5)
)
autoplot(comparison) + ggtitle("Sorting only slows down")
```

There you have it. I decided not to sort. 
